### 2023-09-19 doc/_bayes_freq.qmd
update derivation of bayesian model SI sections

@@

We begin with the probability of the frequency values given the raw barcode
reads. The first assumption is that the inference of the frequency
values for time $t$ is independent of any other time. Therefore, we can write
the joint probability distribution as a product of independent distributions of
the form
$$
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) =
\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t),
$${#eq-freq_indep}
where $\underline{f}_t$ and $\underline{r}_t$ are the $t$-th row of the matrix
containing all of the measurements for time $t$. We imagine that when the 
barcode reads are obtained via sequencing, the quantified number of reads is a
Poisson sample from the "true" underlying number of barcodes within the pool. 
This translates to assuming that the number of reads for each barcode at any
time point $r^{(b)}_t$ is an independent Poisson random variable, i.e.,
$$
r^{(b)}_t \sim \text{Poiss}(\lambda^{(b)}_t),
$${#eq-freq_poisson_reads}
where the symbol "$\sim$" is read "distributed as." Furthermore, for a Poisson
distribution, we have that
$$
\lambda^{(b)}_t = \left\langle r^{(b)}_t \right\rangle = 
\left\langle 
    \left( r^{(b)}_t - \left\langle r^{(b)}_t \right\rangle \right)^2
\right\rangle,
$${#eq-freq_poisson_lambda}
where $\left\langle \cdot \right\rangle$ is the expected value. In other words
the Poisson parameter is equal to the mean and variance of the distribution. The
Poisson distribution has the convenient property that for two Poisson 
distributed random variables $X \sim \text{Poiss}(\lambda_x)$ and 
$Y \sim \text{Poiss}(\lambda_y)$, we have that
$$
Z \equiv X + Y \sim \text{Poiss}(\lambda_x + \lambda_y).
$${#eq-freq_additivity}
This additivity allows us to write the total number of reads at time $t$ $n_t$
also as a Poisson-distributed random variable of the form
$$
n_t \sim \text{Poiss}\left( \sum_{b=1}^B \lambda^{(b)}_t \right),
$${#eq-freq_poisson_total}
where the sum is taken over all $B$ barcodes.

If the total number of reads is given by @eq-freq_poisson_total, the array with
the number of reads for each barcode at time $t$, $\underline{r}_t$ is then
distributed as
$$
\underline{r}_t \sim \text{Multinomial}(n_t, \underline{f}_t),
$${#eq-freq_multinomial}
where each of the $B$ entries of the frequency vector $\underline{f}_t$ is a
function of the $\underline{\lambda}_t$ vector, given by
$$
f_t^{(b)} \equiv f_t^{(b)}(\underline{\lambda}_t) = 
\frac{\lambda_t^{(b)}}{\sum_{b'=1}^B \lambda_t^{(b')}}.
$${#eq-freq_freq_lambda}
In other words, we can think of the $B$ barcode counts as independent Poisson
samples or as a single multinomial draw with a random number of total draws,
$n_t$, and the frequency vector $\underline{f}_t$ we are interested in. Notice
that @eq-freq_freq_lambda is a deterministic function that connects the Poisson
parameters to the frequencies. Therefore, we have the equivalence that
$$
\pi(\underline{f}_t \mid \underline{r}_t) = 
\pi(\underline{\lambda}_t \mid \underline{r}_t),
$${#eq-freq_f_lambda_equiv}
meaning that the uncertainty comes from the $\underline{\lambda}_t$ vector. By
Bayes theorem, we therefore write
$$
\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t),
$${#eq-freq_lambda_bayes}
where we explicitly include the dependence on $n_t$. This does not affect the
distribution or brings more uncertainty because $\underline{r}_t$ already 
contains all the information to compute $n_t$ since
$$
n_t = \sum_{b=1}^B r_t^{(b)}.
$${#eq-freq_n_sum_r}
But adding the variable allows us to factorize @eq-freq_lambda_bayes as
$$
\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) 
\pi(n_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t).
$${#eq-freq_lambda_bayes_factorized}
We then have
$$
\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
\text{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)).
$${#eq-freq_r_bayes}
Furthermore, we have
$$
n_t \mid \underline{\lambda}_t \sim 
\text{Poiss}(\sum_{b=1}^B \lambda_t^{(b)}).
$${#eq=freq_n_bayes}
Finally, for our prior $\pi(\underline{\lambda}_t)$, we first assume each 
parameter is independent, i.e.,
$$
\pi(\underline{\lambda}_t) = \prod_{b=1}^B \pi(\lambda_t^{(b)}).
$$
A reasonable prior for each $\lambda_t^{(b)}$ representing the expected number
of reads for barcode $b$ should span several orders of magnitude. Furthermore,
we assume that no barcode in the dataset ever goes extinct. Thus, no frequency
can equal zero, facilitating the computation of the log frequency ratios
needed to infer the relative fitness. The log-normal distribution satisfies 
these constraints; therefore, for the prior, we assume
$$
\lambda_t^{(b)} \sim 
\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
$$
with $\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}$ as the user-defined 
parameters that characterize the prior distribution.

>>>>

## Defining the Bayesian model

Let us now define each of the terms in @eq-split_posterior described in
@sec-bayesian_inference of the main text. The following sections will specify 
the functional form each of these terms takes.

@@

We begin with the probability of the frequency values given the raw barcode
reads. The first assumption is that the inference of the frequency
values for time $t$ is independent of any other time. Therefore, we can write
the joint probability distribution as a product of independent distributions of
the form
$$
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) =
\prod_{t=1}^T \pi(\underline{f}_t \mid \underline{r}_t),
$${#eq-freq_indep}
where $\underline{f}_t$ and $\underline{r}_t$ are the $t$-th row of the matrix
containing all of the measurements for time $t$. We imagine that when the 
barcode reads are obtained via sequencing, the quantified number of reads is a
Poisson sample from the "true" underlying number of barcodes within the pool. 
This translates to assuming that the number of reads for each barcode at any
time point $r^{(b)}_t$ is an independent Poisson random variable, i.e.,
$$
r^{(b)}_t \sim \operatorname{Poiss}(\lambda^{(b)}_t),
$${#eq-freq_poisson_reads}
where the symbol "$\sim$" is read "distributed as." Furthermore, for a Poisson
distribution, we have that
$$
\lambda^{(b)}_t = \left\langle r^{(b)}_t \right\rangle = 
\left\langle 
    \left( r^{(b)}_t - \left\langle r^{(b)}_t \right\rangle \right)^2
\right\rangle,
$${#eq-freq_poisson_lambda}
where $\left\langle \cdot \right\rangle$ is the expected value. In other words
the Poisson parameter is equal to the mean and variance of the distribution. The
Poisson distribution has the convenient property that for two Poisson 
distributed random variables $X \sim \operatorname{Poiss}(\lambda_x)$ and 
$Y \sim \operatorname{Poiss}(\lambda_y)$, we have that
$$
Z \equiv X + Y \sim \operatorname{Poiss}(\lambda_x + \lambda_y).
$${#eq-freq_additivity}
This additivity allows us to write the total number of reads at time $t$ $n_t$
also as a Poisson-distributed random variable of the form
$$
n_t \sim \operatorname{Poiss}\left( \sum_{b=1}^B \lambda^{(b)}_t \right),
$${#eq-freq_poisson_total}
where the sum is taken over all $B$ barcodes.

If the total number of reads is given by @eq-freq_poisson_total, the array with
the number of reads for each barcode at time $t$, $\underline{r}_t$ is then
distributed as
$$
\underline{r}_t \sim \operatorname{Multinomial}(n_t, \underline{f}_t),
$${#eq-freq_multinomial}
where each of the $B$ entries of the frequency vector $\underline{f}_t$ is a
function of the $\underline{\lambda}_t$ vector, given by
$$
f_t^{(b)} \equiv f_t^{(b)}(\underline{\lambda}_t) = 
\frac{\lambda_t^{(b)}}{\sum_{b'=1}^B \lambda_t^{(b')}}.
$${#eq-freq_freq_lambda}
In other words, we can think of the $B$ barcode counts as independent Poisson
samples or as a single multinomial draw with a random number of total draws,
$n_t$, and the frequency vector $\underline{f}_t$ we are interested in. Notice
that @eq-freq_freq_lambda is a deterministic function that connects the Poisson
parameters to the frequencies. Therefore, we have the equivalence that
$$
\pi(\underline{f}_t \mid \underline{r}_t) = 
\pi(\underline{\lambda}_t \mid \underline{r}_t),
$${#eq-freq_f_lambda_equiv}
meaning that the uncertainty comes from the $\underline{\lambda}_t$ vector. By
Bayes theorem, we therefore write
$$
\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
\pi(n_t, \underline{r}_t \mid \underline{\lambda}_t) \pi(\underline{\lambda}_t),
$${#eq-freq_lambda_bayes}
where we explicitly include the dependence on $n_t$. This does not affect the
distribution or brings more uncertainty because $\underline{r}_t$ already 
contains all the information to compute $n_t$ since
$$
n_t = \sum_{b=1}^B r_t^{(b)}.
$${#eq-freq_n_sum_r}
But adding the variable allows us to factorize @eq-freq_lambda_bayes as
$$
\pi(\underline{\lambda}_t \mid n_t, \underline{r}_t) \propto
\pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
\pi(n_t \mid \underline{\lambda}_t)
\pi(\underline{\lambda}_t)
$${#eq-freq_lambda_bayes_factorized}
We then have
$$
\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)).
$${#eq-freq_r_bayes}
Furthermore, we have
$$
n_t \mid \underline{\lambda}_t \sim 
\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
$${#eq=freq_n_bayes}
Finally, for our prior $\pi(\underline{\lambda}_t)$, we first assume each 
parameter is independent, i.e.,
$$
\pi(\underline{\lambda}_t) = \prod_{b=1}^B \pi(\lambda_t^{(b)}).
$$
A reasonable prior for each $\lambda_t^{(b)}$ representing the expected number
of reads for barcode $b$ should span several orders of magnitude. Furthermore,
we assume that no barcode in the dataset ever goes extinct. Thus, no frequency
can equal zero, facilitating the computation of the log frequency ratios
needed to infer the relative fitness. The log-normal distribution satisfies 
these constraints; therefore, for the prior, we assume
$$
\lambda_t^{(b)} \sim 
\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
$${#eq-freq_lambda_prior}
with $\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}$ as the user-defined 
parameters that characterize the prior distribution.

#### Summary

Putting all the pieces developed in this section together gives a term for our
inference of the form
$$
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}) \propto
\prod_{t=1}^T\left\{
    \pi(\underline{r}_t \mid n_t, \underline{\lambda}_t)
    \pi(n_t \mid \underline{\lambda}_t)
    \left[ 
        \prod_{b=1}^B \pi(\lambda_t^{(b)})
    \right]
\right\}
$${#eq-freq_final_01}
where
$$
\underline{r}_t \mid n_t, \underline{\lambda}_t \sim
\operatorname{Multinomial}(n_t, \underline{f}_t(\underline{\lambda}_t)),
$${#eq-freq_final_02}
$$
n_t \mid \underline{\lambda}_t \sim 
\operatorname{Poiss}\left(\sum_{b=1}^B \lambda_t^{(b)}\right).
$${#eq-freq_final_03}
and
$$
\lambda_t^{(b)} \sim 
\log\mathcal{N}(\mu_{\lambda_t^{(b)}}, \sigma_{\lambda_t^{(b)}}),
$${#eq-freq_final_04}

### 2023-09-19 doc/_bayes_meanfit.qmd
update derivation of bayesian model SI sections

Now, we can focus on one of the terms on the right-hand side of
@eq-meanfit_indep_lineages. Writing Bayes theorem results in
$$
\pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) \pi(\bar{s}_t).
$${#eq-meanfit_bayes}
Notice the likelihood defines the joint distribution of neutral barcode
frequencies conditioned on the population mean fitness. However, rewriting our
fitness model in @eq-fitness for a neutral lineage to leave frequencies on one
side and fitness on the other results in
$$
\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t}\tau.
$${#eq-fitness_ratio_neutral}
@eq-fitness_ratio_neutral implies that our fitness model only relates **the
ratio** of frequencies and not the individual values. To get around this 
complication, we define
$$
\gamma_t^{(b)} \equiv \frac{f_{t+1}^{(b)}}{f_t^{(b)}},
$${#eq-gamma_def}
as the ratio of frequencies between two adjacent time points for any barcode
$b$. This allows us to rewrite the joint distribution
$\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t)$ as
$$
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t).
$${#eq-joint_freq_gamma}
Let us rephrase this subtle but necessary change of variables since it is a key
part of the inference problem: our series of independence assumptions lead us to
@eq-meanfit_bayes that relates the value of the population mean fitness
$\bar{s}_t$ to the frequency of a neutral barcode at times $t$ and $t+1$. 
However, as shown in @eq-fitness_ratio_neutral, our model functionally relates
the ratio of frequencies---that we defined as $\gamma_t^{(n)}$---and not the
independent frequencies to the mean fitness. Therefore, instead of writing for
the likelihood the joint distribution of the frequency values at times $t$ and
$t+1$ conditioned on the mean fitness, we write the joint distribution of the
barcode frequency at time $t$ and the ratio of the frequencies. These **must
be** equivalent joint distributions since there is a one-to-one mapping between
$\gamma_t^{(n)}$ and $f_{t+1}^{(n)}$ for a given value of $f_t^{(n)}$. Another
way to phrase this is to say that knowing the frequency at time $t$ and at time
$t+1$ provides the same amount of information as knowing the frequency at time
$t$ and the ratio of the frequencies. This is because if we want to obtain
$f_{t+1}^{(n)}$ given this information, we simply compute
$$
f_{t+1}^{(n)} = \gamma_t^{(n)} f_t^{(n)}.
$${#eq-meanfit_ft+1_gamma}

The real advantage of rewriting the joint distribution as in
@eq-joint_freq_gamma comes from splitting this joint distribution as a product
of conditional distributions of the form
$$
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)
\pi(\gamma_{t}^{(n)} \mid \bar{s}_t).
$${#eq-joint_to_prod_gamma}
Written in this form, we can finally propose a probabilistic model for how the
mean fitness relates to the frequency ratios we determine in our experiments.
The second term on the right-hand side of @eq-joint_to_prod_gamma relates how
the determined frequency ratio $\gamma_t^{(b)}$ relates to the mean fitness
$\bar{s}_t$. From @eq-fitness_ratio_neutral and @eq-gamma_def, we can write
$$
\ln \gamma_t^{(n)} = - \bar{s}_t + \varepsilon_t^{(n)},
$${#eq-log_ratio_neutral}
where, for simplicity, we set $\tau = 1$. Note that we added an extra term,
$\varepsilon_t^{(n)}$, characterizing the deviations of the measurements from
the theoretical model. It is common to assume that these errors are normally
distributed with mean zero and some standard deviation $\sigma_t$, implying that
$$
\ln \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\mathcal{N}\left(-\bar{s}_t, \sigma_t \right),
$${#eq-log_gamma_normal}
where we include the nuisance parameter $\sigma_t$ to be determined. If we 
assume the log frequency ratio is normally distributed, this implies the 
frequency ratio itself is distributed log-normal. This means that
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\log \mathcal{N}\left(-\bar{s}_t, \sigma_t \right).
$${#eq-meanfit_likelihood}
Having added the nuisance parameter $\sigma_t$ implies that we must update
@eq-meanfit_bayes to
$$
\pi(\bar{s}_t, \sigma_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t) 
\pi(\bar{s}_t) \pi(\sigma_t),
$${#eq-meanfit_bayes_full}
where we assume the prior for each parameter is independent, i.e.,
$$
\pi(\bar{s}_t, \sigma_t) = \pi(\bar{s}_t) \pi(\sigma_t).
$${#eq-meanfit_indep_prior}
For numerical stability, we will select weakly-informative priors for both of
these parameters. In the case of the nuisance parameter $\sigma_t$, the prior
must be restricted to positive values only, since standard deviations cannot be
negative.

@@

Putting all the pieces we have developed in this section together results in an
inference for the population mean fitness values of the form
$$
\pi(
    \underline{\bar{s}}_T, \underline{\sigma}_T \mid \underline{\underline{F}}
) \propto
\prod_{t=1}^{T-1} \left\{
    \prod_{n=1}^N \left[
        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
    \right]
    \pi(\bar{s}_t) \pi(\sigma_t)
\right\},
$${#eq-meanfit_full_inference}
where we have
$$
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
$${#eq-meanfit_eq1}
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
\log\mathcal{N}(\bar{s}_t, \sigma_t),
$${#eq-meanfit_eq2}
$$
\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
$${#eq-meanfit_eq3}
and
$$
\sigma_t \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma_t}),
$${#eq-meanfit_eq4}
where $\sigma_{\bar{s}_t}$ and $\sigma_{\sigma_t}$ are free parameters we can
choose.

>>>>

Now, we can focus on one of the terms on the right-hand side of
@eq-meanfit_indep_lineages. Writing Bayes theorem results in
$$
\pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) \pi(\bar{s}_t).
$${#eq-meanfit_bayes}
Notice the likelihood defines the joint distribution of neutral barcode
frequencies conditioned on the population mean fitness. However, rewriting our
fitness model in @eq-fitness for a neutral lineage to leave frequencies on one
side and fitness on the other results in
$$
\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t\tau}.
$${#eq-fitness_ratio_neutral}
@eq-fitness_ratio_neutral implies that our fitness model only relates **the
ratio** of frequencies and not the individual values. To get around this 
complication, we define
$$
\gamma_t^{(b)} \equiv \frac{f_{t+1}^{(b)}}{f_t^{(b)}},
$${#eq-gamma_def}
as the ratio of frequencies between two adjacent time points for any barcode
$b$. This allows us to rewrite the joint distribution
$\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t)$ as
$$
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t).
$${#eq-joint_freq_gamma}
Let us rephrase this subtle but necessary change of variables since it is a key
part of the inference problem: our series of independence assumptions lead us to
@eq-meanfit_bayes that relates the value of the population mean fitness
$\bar{s}_t$ to the frequency of a neutral barcode at times $t$ and $t+1$. 
However, as shown in @eq-fitness_ratio_neutral, our model functionally relates
the ratio of frequencies---that we defined as $\gamma_t^{(n)}$---and not the
independent frequencies to the mean fitness. Therefore, instead of writing for
the likelihood the joint distribution of the frequency values at times $t$ and
$t+1$ conditioned on the mean fitness, we write the joint distribution of the
barcode frequency at time $t$ and the ratio of the frequencies. These **must
be** equivalent joint distributions since there is a one-to-one mapping between
$\gamma_t^{(n)}$ and $f_{t+1}^{(n)}$ for a given value of $f_t^{(n)}$. Another
way to phrase this is to say that knowing the frequency at time $t$ and at time
$t+1$ provides the same amount of information as knowing the frequency at time
$t$ and the ratio of the frequencies. This is because if we want to obtain
$f_{t+1}^{(n)}$ given this information, we simply compute
$$
f_{t+1}^{(n)} = \gamma_t^{(n)} f_t^{(n)}.
$${#eq-meanfit_ft+1_gamma}

The real advantage of rewriting the joint distribution as in
@eq-joint_freq_gamma comes from splitting this joint distribution as a product
of conditional distributions of the form
$$
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)
\pi(\gamma_{t}^{(n)} \mid \bar{s}_t).
$${#eq-joint_to_prod_gamma}
Written in this form, we can finally propose a probabilistic model for how the
mean fitness relates to the frequency ratios we determine in our experiments.
The second term on the right-hand side of @eq-joint_to_prod_gamma relates how
the determined frequency ratio $\gamma_t^{(b)}$ relates to the mean fitness
$\bar{s}_t$. From @eq-fitness_ratio_neutral and @eq-gamma_def, we can write
$$
\ln \gamma_t^{(n)} = - \bar{s}_t + \varepsilon_t^{(n)},
$${#eq-log_ratio_neutral}
where, for simplicity, we set $\tau = 1$. Note that we added an extra term,
$\varepsilon_t^{(n)}$, characterizing the deviations of the measurements from
the theoretical model. We assume these errors are normally distributed with mean
zero and some standard deviation $\sigma_t$, implying that
$$
\ln \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\mathcal{N}\left(-\bar{s}_t, \sigma_t \right),
$${#eq-log_gamma_normal}
where we include the nuisance parameter $\sigma_t$ to be determined. If we 
assume the log frequency ratio is normally distributed, this implies the 
frequency ratio itself is distributed log-normal. This means that
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\log \mathcal{N}\left(-\bar{s}_t, \sigma_t \right).
$${#eq-meanfit_likelihood}
Having added the nuisance parameter $\sigma_t$ implies that we must update
@eq-meanfit_bayes to
$$
\pi(\bar{s}_t, \sigma_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t) 
\pi(\bar{s}_t) \pi(\sigma_t),
$${#eq-meanfit_bayes_full}
where we assume the prior for each parameter is independent, i.e.,
$$
\pi(\bar{s}_t, \sigma_t) = \pi(\bar{s}_t) \pi(\sigma_t).
$${#eq-meanfit_indep_prior}
For numerical stability, we will select weakly-informative priors for both of
these parameters. In the case of the nuisance parameter $\sigma_t$, the prior
must be restricted to positive values only, since standard deviations cannot be
negative.

@@

Putting all the pieces we have developed in this section together results in an
inference for the population mean fitness values of the form
$$
\pi(
    \underline{\bar{s}}_T, \underline{\sigma}_T \mid \underline{\underline{F}}
) \propto
\prod_{t=1}^{T-1} \left\{
    \prod_{n=1}^N \left[
        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
    \right]
    \pi(\bar{s}_t) \pi(\sigma_t)
\right\},
$${#eq-meanfit_full_inference}
where we have
$$
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
$${#eq-meanfit_eq1}
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
\log\mathcal{N}(\bar{s}_t, \sigma_t),
$${#eq-meanfit_eq2}
$$
\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
$${#eq-meanfit_eq3}
and
$$
\sigma_t \sim \log\mathcal{N}(\mu_{\sigma_t}, \sigma_{\sigma_t}),
$${#eq-meanfit_eq4}
where $\sigma_{\bar{s}_t}$, $\mu_{\sigma_t}$, and $\sigma_{\sigma_t}$ are
user-defined parameters.

### 2023-09-19 doc/_bayes_mutfit.qmd
update derivation of bayesian model SI sections

With all pieces in place, we write the full inference of the relative fitness
values as
$$
\pi(
    \underline{s}^M ,\underline{\sigma}^M \mid 
    \underline{\bar{s}}_T, \underline{\underline{F}}
) \propto
\prod_{m=1}^M \left\{ 
    \pi(f_0^{(m)} \mid \gamma_0^{(m)})
    \prod_{t=0}^{T-1} \left[
        \pi(\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)})
    \right]
    \pi(s^{(m)}) \pi(\sigma^{(m)})
\right\},
$${#eq-mutfit_full_inference}
where
$$
f_0^{(m)} \mid \gamma_0^{(m)} \sim 
\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
$${#eq-mutfit_eq1}
$$
\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
$${#eq-mutfit_eq2}
$$
s^{(m)} \sim \mathcal{N}(0, \sigma_{s^{(m)}}),
$${#eq-mutfit_eq3}
and
$$
\sigma^{(m)} \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma^{(m)}}),
$${#eq-mutfit_eq4}
where $\sigma_{s^{(m)}}$ and $\sigma_{\sigma^{(m)}}$ are free parameters we can
choose.

>>>>

With all pieces in place, we write the full inference of the relative fitness
values as
$$
\pi(
    \underline{s}^M ,\underline{\sigma}^M \mid 
    \underline{\bar{s}}_T, \underline{\underline{F}}
) \propto
\prod_{m=1}^M \left\{ 
    \pi(f_0^{(m)} \mid \gamma_0^{(m)})
    \prod_{t=0}^{T-1} \left[
        \pi(\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)})
    \right]
    \pi(s^{(m)}) \pi(\sigma^{(m)})
\right\},
$${#eq-mutfit_full_inference}
where
$$
f_0^{(m)} \mid \gamma_0^{(m)} \sim 
\operatorname{Uniform}\left(0, \frac{1}{\gamma_0^{(m)}} \right),
$${#eq-mutfit_eq1}
$$
\gamma_t^{(m)} \mid \bar{s}_t, s^{(m)}, \sigma^{(m)} \sim 
\log\mathcal{N}\left(s^{(m)} - \bar{s}_t, \sigma^{(m)} \right),
$${#eq-mutfit_eq2}
$$
s^{(m)} \sim \mathcal{N}(0, \sigma_{s^{(m)}}),
$${#eq-mutfit_eq3}
and
$$
\sigma^{(m)} \sim \log\mathcal{N}(\mu_{\sigma^{(m)}}, \sigma_{\sigma^{(m)}}),
$${#eq-mutfit_eq4}
where $\sigma_{s^{(m)}}$, $\mu_{\sigma^{(m)}}$, and $\sigma_{\sigma^{(m)}}$ are
user-defined parameters.
